Environmental Context Dependency in Species Interactions

### 2018-07-23 EDM_SNI_analysis.Rmd
complete rough draft manuscript

#plot to file if uncommented
# pdf("plots/ccm_network.pdf",height=7,width=7)
plot(g2,asp=0.8, vertex.label.color="black",vertex.label.family="Times", vertex.label.cex=0.6, edge.arrow.size=0.8,edge.arrow.width=0.8,vertex.label.font=2, vertex.shape="circle", edge.curved=curves,margin=0,vertex.size=20,rescale=T,edge.lty=1,edge.color=edge.cols,layout=layout_in_circle,vertex.color=vert.cols,vertex.frame.color="black",main="")
# dev.off()
```

@@

@@

```{r interactions density plot by type,fig.height=6,fig.width=6}
# Density plot
interaction_distribution_by_type <- spp.only.mod.coeffs %>%
  filter(!is.na(type),type != "intraspecies")%>%
  group_by(type) %>%
  mutate(mean_int=mean(value,na.rm=T))%>%
  ungroup()%>%
  ggplot(aes(value,..density..))+
  geom_density(aes(fill=type),alpha=0.8,col=NA)+
  geom_vline(xintercept=0,linetype=2)+
  geom_vline(aes(xintercept=mean_int))+
  scale_fill_manual(values=c("darkorange1","orangered2","darkorchid1","palegreen3","skyblue1"),name="")+
   guides(fill=F)+
  xlim(-1.5,1.5)+
  labs(x="Interaction strength",y="Kernel density")+
  geom_text(aes(x=-1.5,y=0.5,label=type),hjust="inward",family="Rockwell",fontface="italic",check_overlap = TRUE,size=4,show.legend = FALSE)+
  facet_grid(rows=vars(type))+
  theme(strip.text = element_blank())

@@

>>>>

#plot to file if uncommented
# png("plots/ccm_network.png",height=500,width=600)
plot(g2,asp=0.8, vertex.label.color="black",vertex.label.family="Times", vertex.label.cex=0.6, edge.arrow.size=0.8,edge.arrow.width=0.8,vertex.label.font=2, vertex.shape="circle", edge.curved=curves,margin=0,vertex.size=20,rescale=T,edge.lty=1,edge.color=edge.cols,layout=layout_in_circle,vertex.color=vert.cols,vertex.frame.color="black",main="")
# dev.off()
```

@@

```{r pter effect ymac,fig.height=6,fig.width=6}
pter_effect_ymac <-spp.only.coeffs.with.phys %>%
  ungroup()%>%
  filter(mod=="ymac",effect=="pter")%>%
  mutate(phys_high_low=case_when(
    ext_value > 1 ~ "high",
    ext_value < -1 ~ "low")) %>%
  filter(!is.na(phys_high_low))%>%
  left_join(select(westend,period,one_of(phys.vars)),by="period")%>%
  group_by(type,external_forcing,phys_high_low)%>%
  mutate(mean_int=mean(value,na.rm=T))%>%
  distinct()%>%
  ungroup()

pter_effect_ymac_plot <-pter_effect_ymac %>%
ggplot(aes(value,..density..,fill=phys_high_low))+
  geom_density(col=NA,alpha=0.6)+
  geom_vline(xintercept=0,linetype=2)+
  geom_vline(aes(xintercept=mean_int,color=phys_high_low),size=1.5)+
  scale_fill_manual(values=c("darkorange","dodgerblue4"),name="Physical\nForcing Value",labels=c("high","low"))+
  scale_color_manual(values=c("darkorange","dodgerblue4"))+
  guides(color="none")+
  labs(x="Effect of Pterygophora on Macrocystis Recruits",y="Probability Density")+
  facet_grid(rows=vars(external_forcing),labeller=labeller(external_forcing=phys.labeller))+
  theme(strip.text=element_text(size=10))

# ggsave("plots/pter_effect_ymac_phys.png",plot=pter_effect_ymac_plot,h=6,w=6,dpi=500)
# ggsave("plots/pter_effect_ymac_phys.pdf",plot=pter_effect_ymac_plot,h=6,w=6)
pter_effect_ymac_plot
```

@@

```{r interactions density plot by type,fig.height=6,fig.width=6}
# Density plot
interaction_distribution_by_type <- spp.only.mod.coeffs %>%
  filter(!is.na(type),type != "intraspecies")%>%
  group_by(type) %>%
  mutate(mean_int=mean(value,na.rm=T))%>%
  ungroup()%>%
  ggplot(aes(value,..density..))+
  geom_density(aes(fill=type),alpha=0.8,col=NA)+
  geom_vline(xintercept=0,linetype=2)+
  geom_vline(aes(xintercept=mean_int))+
  scale_fill_manual(values=c("darkorange1","orangered2","darkorchid1","palegreen3","skyblue1"),name="")+
   guides(fill=F)+
  xlim(-1.5,1.5)+
  labs(x="Interaction strength",y="Probability density")+
  geom_text(aes(x=-1.5,y=0.5,label=type),hjust="inward",family="Rockwell",fontface="italic",check_overlap = TRUE,size=4,show.legend = FALSE)+
  facet_grid(rows=vars(type))+
  theme(strip.text = element_blank())

@@

# Mean vs. range (95%)

int_mean_range_plot <- int_mean_sd %>%
  filter(type != "intraspecies", effect != "ymac", mod != "ymac")%>%
  ggplot(aes(abs_mean,y=upper-lower))+
  geom_point(aes(color=type,shape=type,fill=type),size=5)+
  geom_smooth(method="lm",se=F,col="black",linetype=2,show.legend = F)+
  # geom_vline(xintercept=0,linetype=2)+
  # geom_encircle(aes(fill=type),col=NA,alpha=0.3,expand=0.1)+
  scale_color_manual(values=c("darkorange1","orangered2","darkorchid1","palegreen3","skyblue1"),name="")+
  scale_fill_manual(values=c("darkorange1","orangered2","darkorchid1","palegreen3","skyblue1"),name="")+
  scale_shape_manual(values=c(15,16,17,18),name="")+
  guides(fill='none')+
  labs(x="Mean Interaction Strength",y="Range Interaction Strength")+
  theme(legend.position = c(0.8,0.8),
        legend.background = element_rect(color="black"))
# ggsave("plots/int_mean_range.png",plot=int_mean_range_plot,h=5,w=6)
# ggsave("plots/int_mean_range.pdf",plot=int_mean_sd_plot)
int_mean_range_plot

### 2018-07-23 manuscript.Rmd
complete rough draft manuscript

---
title: "Environmental Context Dependency in Species Interactions"
author: "Owen Liu"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
fontsize: 12pt
bibliography: citation_bibtex.bib
csl: nature.csl
editor_options: 
 chunk_output_type: inline
header-includes:
  - \usepackage[left]{lineno}
  - \linenumbers
  - \renewcommand\linenumberfont{\normalfont\bfseries\small}
---

@@

```{r packages and workspace}
require(tidyverse)
require(extrafont)
load("ws/wrkspce_07.19.18.RData")

@@

* Maximum 3500 words main text (excluding introductory pa ragraph/abstract)
* Maximum 6 display items (figures/tables)
* The introductory paragraph is typically 150 words and is unreferenced; it contains a brief account of the background and rationale of the work, followed by a statement of the main conclusions introduced by the phrase "Here we show" or some equivalent.An introduction (without heading) of up to 500 words of referenced text expands on the background of the work (some overlap with the summary is acceptable), and is followed by a concise, focused account of the findings (headed 'Results'), and one or two short paragraphs of discussion (headed 'Discussion').

@@

**FIGURES**

Time series of study species
Example attractor? (in supplementary)
CCM results

Ecological interactions are not uniform across time, and instead vary with environmental conditions. Interactions among species are often measured with short-term controlled experiments, but these experiments are subject to the particular environmental conditions under which they are performed. As an alternative, we utilize empirical dynamic modeling applied to a 30-year time series to estimate species interaction strengths across a wide range of environmental conditions in a coastal marine ecosystem. By including large-scale climate indices, sea surface temperature, and a measure of physical disturbance in analyses, we show that environmental context influences the strength and direction of species interactions. In so doing, we are able to confirm and extend results from previous studies, as well as identify potentially important but understudied interactions. The significant context dependency in species interactions found in this study argues for a greater utilization of long-term data and empirical dynamic modeling in studies of ecosystem dynamics.

A solution to these difficulties is to a) utilize ecological observations collected over a long time period, across a large range of environmental contexts, with b) an analytical method to directly estimate context-dependent species interactions from those observations. Such an approach could help to characterize environmental contingencies in species interactions and explicitly examine interaction variability. Here, we use empirical dynamic modelling (EDM, [@Deyle2016]) to estimate a varying species interaction network and establish environmental context dependency in interaction strength and direction. Empirical dynamic modelling uses information from single or multiple time series to empirically model relationships between variables through the reconstruction of dynamic attractors (https://youtu.be/8DikuwwPWsY). The general modelling framework for all EDM methods is readily adaptable to many different sorts of time series variables, including environmental variables manifesting at different scales [@Deyle2013; @Chang2017; @Ye2015]. Because the methods are specifically designed for nonlinear dynamic systems, EDM---in theory---should be able to illuminate context-dependent patterns in species interactions.

## Results and discussion

```{r fig.width=4, fig.height=4,fig.cap="Reconstructed interaction web using results of convergent cross mapping\\label{fig:ccm}"}
include_graphics("plots/ccm_network.pdf")
```

```{r fig.cap="Distribution of estimated interaction by type\\label{fig:int_dist}"}
interaction_distribution_by_type
```

In the San Nicolas kelp forest, we find a striking prevalence of neutral and positive species interactions (Figure \ref{fig:int_dist}). After grouping species interactions by type, only herbivory (the effect of urchins on algae species) is predominantly negative. Conversely, the effect of the algae species on the urchins has the flattest distribution, with occasional strong negative and strong positive interactions. Likewise, contrary to our expectations, interactions between the algae species and between the urchin species are not always antagonistic. These results suggest that facilitation is common in kelp forests when context is considered. Since there is evidence here of strong herbivory, for example, then the indirect facilitative effects between algae species may sometimes outweigh their direct competition. Additionally, *Macrocystis* forests can mediate current strength and as a result, can also help retain reproductive propagules of other species[@Graham2007c]. These types of indirect facilitation in kelp forests have received little attention[@Arkema2009a; Benes2015] relative to a strong focus on exploitative competition between these species for light and nutrients[@Reed1984; @Dayton1984], but similar effects have been documented in other systems[@Baraza2006]. Our analysis does not contradict the importance of competition in kelp forest ecosystems, but suggests that facilitation, especially indirect facilitation, may be an important structuring factor in kelp forest ecosystems.

```{r fig.cap="Mean and standard deviation of each estimated unidirectional species interaction"}
int_mean_sd_plot
```

```{r fig.cap="Macrocystis effect on Pterygophora"}
mac_effect_on_pter_plot
```

>>>>

---
title: "Environmental Context Dependency in Species Interactions"
author: "Owen Liu"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  word_document: default
  pdf_document: default
editor_options:
  chunk_output_type: inline
fontsize: 12pt
header-includes:
- \usepackage[left]{lineno}
- \linenumbers
- \renewcommand\linenumberfont{\normalfont\bfseries\small}
csl: nature.csl
bibliography: citation_bibtex.bib
---

@@

```{r packages and workspace}
# # header-includes:
#   - \usepackage[left]{lineno}
#   - \linenumbers
#   - \renewcommand\linenumberfont{\normalfont\bfseries\small}
require(tidyverse)
require(extrafont)
load("ws/wrkspce_07.20.18.RData")

@@

* Maximum 3500 words main text (excluding introductory paragraph/abstract)
* Maximum 6 display items (figures/tables)
* The introductory paragraph is typically 150 words and is unreferenced; it contains a brief account of the background and rationale of the work, followed by a statement of the main conclusions introduced by the phrase "Here we show" or some equivalent.An introduction (without heading) of up to 500 words of referenced text expands on the background of the work (some overlap with the summary is acceptable), and is followed by a concise, focused account of the findings (headed 'Results'), and one or two short paragraphs of discussion (headed 'Discussion').

Ecological interactions are not uniform across time, and instead vary with environmental conditions. Interactions among species are often measured with short-term controlled experiments, but these experiments are subject to the particular environmental conditions under which they are performed. As an alternative, we utilize empirical dynamic modeling applied to a 30-year time series to estimate species interactions across a wide range of environmental conditions. We show that environmental context influences the strength and direction of species interactions. In so doing, we are able to confirm and extend results from previous studies, as well as identify potentially important but understudied dynamics, including the importance of specifically studying variation rather than mean interaction outcomes. The significant context dependency in species interactions found in this study argues for a greater utilization of long-term data and empirical dynamic modeling in studies of ecosystem dynamics.

A solution to these difficulties is to a) utilize ecological observations collected over a long time period, across a large range of environmental contexts, with b) an analytical method to directly estimate context-dependent species interactions from those observations. Such an approach could help to characterize environmental contingencies in species interactions and explicitly examine interaction variability. Here, we use empirical dynamic modelling (EDM[@Deyle2016]) to estimate a varying species interaction network and establish environmental context dependency in interaction strength and direction. Empirical dynamic modelling uses information from single or multiple time series to empirically model relationships between variables through the reconstruction of dynamic attractors (https://youtu.be/8DikuwwPWsY). The general modelling framework for all EDM methods is readily adaptable to many different sorts of time series variables, including environmental variables manifesting at different scales [@Deyle2013; @Chang2017; @Ye2015]. Because the methods are specifically designed for nonlinear dynamic systems, EDM---in theory---should be able to illuminate context-dependent patterns in species interactions.

## Results

```{r fig.width=4, fig.height=4,fig.cap="Reconstructed interaction web using results of convergent cross mapping. Each arrow represents a significant causal signal, and link width and opacity scale with the strength of causal forcing (see full results in Supplementary material). Species abbreviations: M. pyr: \\textit{Macrocystis pyrifera}; L.far: \\textit{Laminaria farlowii}; P.cal: \\textit{Pterygophora californica}; M.fra: \\textit{Mesocentrotus franciscanus}; S.pur: \\textit{Strongylocentrotus purpuratus}. Physical drivers: NPGO: North Pacific Gyre Oscillation; MEI: Multivariate El Nino Index; PDO: Pacific Decadal Oscillation; SST: Sea surface temperature; SWH: Significant wave height.\\label{fig:ccm}"}
include_graphics("plots/ccm_network.pdf")
```

```{r fig.cap="Smoothed kernel density histograms of estimated interactions by type. Solid lines denote means across all estimated interactions of that type.\\label{fig:int_dist}"}
interaction_distribution_by_type
```

In the San Nicolas kelp forest, we find a striking prevalence of neutral and positive species interactions (Figure \ref{fig:int_dist}). After grouping species interactions by type, only herbivory (the effect of urchins on algae species) is predominantly negative. Conversely, the effect of the algae species on the urchins has the flattest distribution, with occasional strong negative and strong positive interactions. Likewise, contrary to our expectations, interactions between the algae species and between the urchin species are not always antagonistic. These results suggest that facilitation---direct or indirect---can arise in kelp forests in multiple contexts[@Bennett2015]. Since there is evidence here of strong herbivory, for example, then the indirect facilitative effects between algae species through shared herbivory avoidance[@Harris1984] may sometimes outweigh their direct competition. Additionally, *Macrocystis* forests can mediate current strength and as a result, can also help retain reproductive propagules of other species[@Graham2007c]. These types of indirect facilitation in kelp forests have received relatively little attention[@Arkema2009a; @Benes2015] relative to a strong focus on exploitative competition between these species for light and nutrients[@Reed1984; @Dayton1984], but similar effects have been documented in other systems[@Baraza2006]. Our analysis does not contradict the importance of competition in kelp forest ecosystems, but suggests that facilitation, especially indirect facilitation, may be an additional important structuring force.

```{r fig.cap="Mean and standard deviation of each estimated unidirectional species interaction (excluding intraspecies effects)\\label{fig:int_mean_range}"}
int_mean_range_plot
```

There is an inverse relationship between mean and range in interaction strengths (Figure \ref{fig:int_mean_range}). Importantly, these reported ranges in interaction strength are distinct from estimation or experimental errors. They represent observed ranges across all individually-estimated S-map interactions from each reconstructed attractor.  While inverse mean-range relationship was true overall, it varied by interaction type. For the three significant interactions between algae species (excluding interactions with *Macrocytis* recruits), the effects with the smallest mean strength had the largest range across the time series. The effects of algal on urchin species were the most variable in general, and the herbivory effects the least. Additionally, the range was greater than the mean for all of these interactions, further supporting the idea that context dependency may be the rule rather than the exception in this ecosystem. The observed mean-range relationship echoes work done across a range of experimental systems[@Berlow1999b], and implies that interactions that seem weak on average may simply vary in sign and magnitude across contexts but still be important in each of those contexts nonetheless.

```{r fig.cap="Distributions of \\textit{Macrocystis} effects on \\textit{Pterygophora}, under high (greater than 1) and low (less than -1) values of five normalized environmental indices (abbreviations as in Figure \\ref{fig:ccm}. Solid lines: mean interactions under each regime.\\label{fig:mac_pter}"}
mac_effect_on_pter_plot
```

As a case in point, consider the interaction of adult *Macrocystis* and the understory kelp *Pterygophora californica*,  the interaction with the "weakest" mean strength of all algal interactions in Figure \ref{fig:int_mean_range}. *Macrocystis* is often the dominant competitor in kelp forests for nutrients and light[@Dayton1984; @Reed1984], but in this site on average has a neutral to positive effect on *Pterygophora* (Figure \ref{fig:mac_pter}), a result seemingly incongruous with the established competitive hierarchy. However, in southern California, *Macrocystis* is also known to be a better competitor under cold-water, nutrient-rich conditions[@Dayton1999]. Although negative interactions of *Macrocystis* with *Pterygophora* are rare in our study, they align with those expectations: *Macrocystis* has a greater chance of flipping its interaction from positive to negative---and asserting interspecies dominance---under low values of the MEI and PDO, and high values of the NPGO (Figure \ref{fig:mac_pter}). These are all climate regimes associated with increased availability of nutrients in southern California[@Mantua2002; @Wolter2011; @DiLorenzo2008]. Hence, the rare, but strong negative effects of *Macrocystis* on *Pterygophora* conform to expected patterns across decadal-scale climate shifts[@Dayton1999; @Tegner1997], despite the two species' weak mean interaction at this site. In this way, the prevailing environmental context at any point in time is the landscape upon which species interactions occur, potentially tipping the scales in favor of one species over another. Moreover, these rare competitive events may be precisely what precipitates nonlinear ecosystem shifts between alternative states.

## Discussion

Ecosystem dynamics are composed of nonlinear species relationships, played out within shifting environmental contexts. A significant challenge in the study of ecosystem dynamics has been the difficulty in appropriately extrapolating experimental results to real ecosystems, where multiple species-species and species-environment interactions are operating simulateously. We have shown in this study that empirical dynamic modeling can help to tackle this challenge, using time series data to accurately reconstruct nonlinear ecosystem trajectories. Beginning with a published monitoring data set from a kelp forest ecosystem, EDM methods helped to elucidate causation, build interaction networks, and investigate the influence of large-scale environmental drivers on interaction strength. In this particular ecosystem, our analyses confirmed decades of experimental work regarding the foundation species *Macrocystis pyrifera*, but also were able to contextualize those classic interactions as they apply at the study site. A classic algal competitive dominance hierarchy[@Dayton1999] is seemingly weak at this site under average conditions, but not absent---under predictable nutrient, temperature, and disturbance regimes, *Macrocystis* can be dominant at this site.

More generally, we report evidence here of an inverse mean-range relationship in interaction strengths and a strong presence of apparent facilitation under multiple environmental contexts, results that agree with previous theoretical and experimental findings across multiple ecosystems[@McCann1998; @Berlow1999b; @Bruno2003; @Bulleri2009]. An important implication is that if a goal is understanding the dynamics of entire ecosystems, studying the mean outcome of single species interactions may not be adequate. Indirect associations between multiple species and shifting environmental contexts may give rise to rare, critical moments when fleeting strong interactions determine ecosystem shifts. This idea needs further investigation in multiple systems, but if widely applicable, it means that context dependency---and its role in mediating varying species interaction strengths---deserves more attention than the identification of context-averaged mean interactions[@Chamberlain2014]. In a growing number of ecosystems, EDM is helping in this endeavour[@Deyle2016; @Ye2015a; @McGowan2017].

EDM does not take the place of experimentation, but we argue that it can help to both contextualize and guide it. Our analysis was a proof of concept: we started with simple time series from a monitoring dataset in a well-studied but complex ecosystem, and showed how previous experimental results play out over a longer time period. We hope that this approach can be applied to many other ecosystems, where time series data exist but where important interactions may not be as well-established. Where important interactions are known, EDM can help to explore whether environmental context matters in interaction variance. Where those interactions are not as well known, EDM may be a helpful first step in identification of ecosystem links whose mechanisms can then be further established through other methods.

## Extra notes (scratch space)

The disturbance effect is interesting as well[@Tegner1997]. Large disturbances can rip out entire forests (*Macrocystis* and *Pterygophora* alike) and can be the catalyst for phase shifts in ecosystem state. That is evident here in the negative tail of the effect of *Macrocystis* on *Pterygophora*. If conditions are beneficial for rapid growth, *Macrocystis* can establish more quickly than *Pterygophora* after disturbance, thereby achieving dominance and having a negative effect on *Pterygoph
ora*.

The realized competitive niche of *Macrocystis*[@Dayton1999] may be altered at the study site, which is comprised of a low-rugosity reef highly exposed to wave disturbance, habitat characteristics that can inhibit or even reverse *Macrocystis* dominance[@Graham1997]. The simple observation that the San Nicolas Island site has not maintained a stable kelp forest for significant periods of time further supports the lack of consistent *Macrocystis* dominance.

@@
